

##R Tutorial for Commuting Information with IPUMS USA 

The first step is loading all the required packages. Some may need to be installed - use the code: install.package(name_of_package)

```{r message = FALSE, warning = FALSE}
#loading all the necessary packages library
library(ipumsr)
library(truncnorm)
library(reshape2)
library(ggplot2)
library(tidyverse)
library(qwraps2)
library(knitr)
library(kableExtra)
```


Next, load the data files into RStudio. Both the .dat file and DDI codebook should be saved in the Working Directory.

```{r message = FALSE, warning = FALSE}
ddi <- read_ipums_ddi("usa_00003.xml")
df4 <- read_ipums_micro(ddi)
seadf <- data.frame(df4$AGE, df4$HHINCOME, df4$RENTGRS, df4$TRANWORK, df4$TRANTIME, df4$SEX, df4$METRO, stringsAsFactors = FALSE)
names(seadf) <- c("AGE", "INCOME", "RENT", "TRANWORK", "COMMUTE_TIME", "SEX", "METRO")
```



Many of the variables in the IPUMS data contain missing values or numeric codes for non-responses that will impact analysis. These need to be removed or changed prior to creating summary statistics.
```{r}
#Changing HHINCOME to not replace 9999999 values as NA
seadf <- seadf %>%
 mutate(INCOME = replace(INCOME, INCOME == 9999999, NA))
```

```{r}
#Removing 0 values from rent variable
seadf <- seadf %>%
 mutate(RENT = replace(RENT, RENT == 0, NA))
```


```{r}
#Recoding variable for METRO to correspond with matrix
#See IPUMS definition for METRO at https://usa.ipums.org/usa-action/variables/METRO#description_section

seadf$METRO1 <- NA
seadf$METRO1[seadf$METRO==2] <- "Urban"
seadf$METRO1[seadf$METRO==0 & seadf$METRO==1] <- "Non-Urban"
seadf$METRO1[seadf$METRO==3 & seadf$METRO==4] <- "Mixed"
```


```{r}
#Recoding TRANWORK variable to correspond with matrix 
#See IPUMS definition for TRANWORK at #https://usa.ipums.org/usa-action/variables/TRANWORK#description_section 

seadf$COMMUTE_TYPE <-NA 
seadf$COMMUTE_TYPE[seadf$TRANWORK>=10 & seadf$TRANWORK<=20] <- "Private Motorized Vehicle" 
seadf$COMMUTE_TYPE[seadf$TRANWORK>=30 & seadf$TRANWORK<=36] <- "Public Transit" 
seadf$COMMUTE_TYPE[seadf$TRANWORK>=40 & seadf$TRANWORK<=50] <- "Bike/Walk/Other"  
```

```{r}
#Dropping all observations that did not commute
seadf <- drop_na(seadf, COMMUTE_TYPE)
```


Once the data has beeen prepared for analysis, the next step is to create a table of summary statistics for each cell in the matrix. The first table in this example includes information for the entire Seattle data set. Each subsequent table creates a subsetted dataframe based on METRO and provides summary statistics for the row in the matrix. 

```{r message = FALSE, warning = FALSE}
#Creating a table of summary statistics for all the data

options(qwraps2_markup = "markdown")

#Creating a table of summary statistics
sumtable1 <- 
  list("Age" =
         list('Age: mean (sd)'=~qwraps2::mean_sd(AGE, denote_sd='paren')),
       "Income" = 
         list('Income: mean'=~mean(INCOME, na.rm=TRUE)),
       "Sex" = 
         list('# Female (%)' = ~ qwraps2::n_perc0(SEX==2),
              '# Male (%)' = ~ qwraps2::n_perc0(SEX==1)),
       'Commute Time' =
         list('Average (sd)'=~qwraps2::mean_sd(COMMUTE_TIME, denote_sd='paren'),
              'Median and IQR)'=~qwraps2::median_iqr(COMMUTE_TIME)))

#Putting the summary statistics into a dataframe
test <- summary_table(dplyr::group_by(seadf, COMMUTE_TYPE), sumtable1)

#Naming the columns
colnames(test) <- c("Bike/Walk/Other", "Private Motorized Vehicle", "Public Transit")

test %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 13) %>%
  pack_rows("Sex", 3, 4) %>%
  pack_rows("Commute Time", 5, 6)

```





```{r message = FALSE, warning = FALSE}
#Summary statistics for urban commuters
urbandata <- subset(seadf, METRO==2)

sumtable2 <- 
  list("Age" =
         list('Age: mean (sd)'=~qwraps2::mean_sd(AGE, denote_sd='paren')),
       "Income" = 
         list('Income: mean'=~mean(INCOME, na.rm=TRUE)),
       "Sex" = 
         list('# Female (%)' = ~ qwraps2::n_perc0(SEX==2),
              '# Male (%)' = ~ qwraps2::n_perc0(SEX==1)),
       'Commute Time' =
         list('Average (sd)'=~qwraps2::mean_sd(COMMUTE_TIME, denote_sd='paren'),
              'Median and IQR)'=~qwraps2::median_iqr(COMMUTE_TIME)))

#Putting the summary statistics into a dataframe
urban_table <- summary_table(dplyr::group_by(urbandata, COMMUTE_TYPE), sumtable2)

#Naming the columns
colnames(urban_table) <- c("Bike/Walk/Other", "Private Motorized Vehicle", "Public Transit")

urban_table %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 13) %>%
  pack_rows("Sex", 3, 4) %>%
  pack_rows("Commute Time", 5, 6)
```



```{r message = FALSE, warning = FALSE}
#Summary statistics for non-urban commuters

non_urbandata <- subset(seadf, METRO==0 | METRO ==1)

sumtable3 <- 
  list("Age" =
         list('Age: mean (sd)'=~qwraps2::mean_sd(AGE, denote_sd='paren')),
       "Income" = 
         list('Income: mean'=~mean(INCOME, na.rm=TRUE)),
       "Sex" = 
         list('# Female (%)' = ~ qwraps2::n_perc0(SEX==2),
              '# Male (%)' = ~ qwraps2::n_perc0(SEX==1)),
       'Commute Time' =
         list('Average (sd)'=~qwraps2::mean_sd(COMMUTE_TIME, denote_sd='paren'),
              'Median and IQR)'=~qwraps2::median_iqr(COMMUTE_TIME)))

#Putting the summary statistics into a dataframe
non_urban_table <- summary_table(dplyr::group_by(non_urbandata, COMMUTE_TYPE), sumtable2)

#Naming the columns
colnames(non_urban_table) <- c("Bike/Walk/Other", "Private Motorized Vehicle", "Public Transit")

non_urban_table %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 13) %>%
  pack_rows("Sex", 3, 4) %>%
  pack_rows("Commute Time", 5, 6)
```

```{r message = FALSE, warning = FALSE}
#Summary statistics for mixed commuters

mixeddata <- subset(seadf, METRO==3)

sumtable4 <- 
  list("Age" =
         list('Age: mean (sd)'=~qwraps2::mean_sd(AGE, denote_sd='paren')),
       "Income" = 
         list('Income: mean'=~mean(INCOME, na.rm=TRUE)),
       "Sex" = 
         list('# Female (%)' = ~ qwraps2::n_perc0(SEX==2),
              '# Male (%)' = ~ qwraps2::n_perc0(SEX==1)),
       'Commute Time' =
         list('Average (sd)'=~qwraps2::mean_sd(COMMUTE_TIME, denote_sd='paren'),
              'Median and IQR)'=~qwraps2::median_iqr(COMMUTE_TIME)))

#Putting the summary statistics into a dataframe
mixed_table <- summary_table(dplyr::group_by(mixeddata, COMMUTE_TYPE), sumtable2)

#Naming the columns
colnames(mixed_table) <- c("Bike/Walk/Other", "Private Motorized Vehicle", "Public Transit")

mixed_table %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 13) %>%
  pack_rows("Sex", 3, 4) %>%
  pack_rows("Commute Time", 5, 6)
```

After we have obtained summary statistics, we can begin looking at the interaction between commute type, proximity to downtown, and values. Because the indicator for value has not been measured either in this sample or created as an additonal variable through a survey, indicators will be generated randomly in a normal distribution between 0 and 1 (0 being the weakest indicator and 1 being the strongest). 

```{r message = FALSE, warning = FALSE}
#This set of code generates a random indicator for value, it is not a necessary component of the final toolkit 
#find row values to see random seeds
nrow(urbandata)
nrow(non_urbandata)
nrow(mixeddata)
```


```{r}
#setting value variable as normal distribution between 0 and 1
value_urban_hedo <- rtruncnorm(52277, a=0, b=1)
value_urban_bio <- rtruncnorm(52277, a=0, b=1)
value_urban_ego <- rtruncnorm(52277, a=0, b=1)
value_urban_altru <- rtruncnorm(52277, a=0, b=1)

urbandata$value_hedo <- value_urban_hedo
urbandata$value_ego <- value_urban_ego
urbandata$value_bio <- value_urban_bio
urbandata$value_altru <- value_urban_altru

value_nonurban_hedo <- rtruncnorm(69011, a=0, b=1)
value_nonurban_bio <- rtruncnorm(69011, a=0, b=1)
value_nonurban_ego <- rtruncnorm(69011, a=0, b=1)
value_nonurban_altru <- rtruncnorm(69011, a=0, b=1)

non_urbandata$value_hedo <- value_nonurban_hedo
non_urbandata$value_ego <- value_nonurban_ego
non_urbandata$value_bio <- value_nonurban_bio
non_urbandata$value_altru <- value_nonurban_altru

value_mixed_hedo <- rtruncnorm(104665, a=0, b=1)
value_mixed_bio <- rtruncnorm(104665, a=0, b=1)
value_mixed_ego <- rtruncnorm(104665, a=0, b=1)
value_mixed_altru <- rtruncnorm(104665, a=0, b=1)

mixeddata$value_hedo <- value_mixed_hedo
mixeddata$value_ego <- value_mixed_ego
mixeddata$value_bio <- value_mixed_bio
mixeddata$value_altru <- value_mixed_altru
```

Now that we have established value indicators for each observation, we need to find the mean for each cell in the matrix. This will allow us to visualize the values on a radar plot.

```{r message = FALSE, warning = FALSE, results=FALSE}
#mean for urban private motorized vehicle hedo
mean(urbandata$value_hedo[urbandata$TRANWORK>=10 & urbandata$TRANWORK<=20])

#mean for urban public transportation hedo
mean(urbandata$value_hedo[urbandata$TRANWORK>=30 & urbandata$TRANWORK<=36])

#mean for urban bike/walk/other hedo
mean(urbandata$value_hedo[urbandata$TRANWORK>=40 & urbandata$TRANWORK<=60])

#mean for urban private motorized vehicle ego
mean(urbandata$value_ego[urbandata$TRANWORK>=10 & urbandata$TRANWORK<=20])

#mean for urban public transportation ego
mean(urbandata$value_ego[urbandata$TRANWORK>=30 & urbandata$TRANWORK<=36])

#mean for urban bike/walk/other ego
mean(urbandata$value_ego[urbandata$TRANWORK>=40 & urbandata$TRANWORK<=60])

#mean for urban private motorized vehicle altru
mean(urbandata$value_altru[urbandata$TRANWORK>=10 & urbandata$TRANWORK<=20])

#mean for urban public transportation altru
mean(urbandata$value_altru[urbandata$TRANWORK>=30 & urbandata$TRANWORK<=36])

#mean for urban bike/walk/other altru
mean(urbandata$value_altru[urbandata$TRANWORK>=40 & urbandata$TRANWORK<=60])

#mean for urban private motorized vehicle bio
mean(urbandata$value_bio[urbandata$TRANWORK>=10 & urbandata$TRANWORK<=20])

#mean for urban public transportation bio
mean(urbandata$value_bio[urbandata$TRANWORK>=30 & urbandata$TRANWORK<=36])

#mean for urban bike/walk/other bio
mean(urbandata$value_bio[urbandata$TRANWORK>=40 & urbandata$TRANWORK<=60])
```

```{r message = FALSE, warning = FALSE}
#Creating a dataframe with average value indicators

hedo_urban <- c(mean(urbandata$value_hedo[urbandata$TRANWORK>=10 & urbandata$TRANWORK<=20]), mean(urbandata$value_hedo[urbandata$TRANWORK>=30 & urbandata$TRANWORK<=36]), mean(urbandata$value_hedo[urbandata$TRANWORK>=40 & urbandata$TRANWORK<=60]))
ego_urban <- c(mean(urbandata$value_ego[urbandata$TRANWORK>=10 & urbandata$TRANWORK<=20]), mean(urbandata$value_ego[urbandata$TRANWORK>=30 & urbandata$TRANWORK<=36]), mean(urbandata$value_ego[urbandata$TRANWORK>=40 & urbandata$TRANWORK<=60]))
altru_urban <- c(mean(urbandata$value_altru[urbandata$TRANWORK>=10 & urbandata$TRANWORK<=20]), mean(urbandata$value_altru[urbandata$TRANWORK>=30 & urbandata$TRANWORK<=36]), mean(urbandata$value_altru[urbandata$TRANWORK>=40 & urbandata$TRANWORK<=60]))
bio_urban <- c(mean(urbandata$value_bio[urbandata$TRANWORK>=10 & urbandata$TRANWORK<=20]), mean(urbandata$value_bio[urbandata$TRANWORK>=30 & urbandata$TRANWORK<=36]), mean(urbandata$value_bio[urbandata$TRANWORK>=40 & urbandata$TRANWORK<=60]))
group <- c("Private Motorized Vehicle", "Public Transportation", "Bike/Walk/Other")

radar_urban <- data.frame(group, hedo_urban, ego_urban, altru_urban, bio_urban)


```

```{r}
#creating a new datafram with set meanavlues for radar plots

hedo_urban <- c(0.52, 0.41, .35)
ego_urban <- c(0.42, 0.32, 0.34)
altru_urban <- c(0.3, 0.52, 0.54)
bio_urban <- c(0.41, 0.56, 0.58)

radar_urban_test <- data.frame(group, hedo_urban, ego_urban, altru_urban, bio_urban)

```


```{r}
View(radar_urban)
```


```{r message = FALSE, warning = FALSE}
#Setting up the data for ggplot
urban_toplot=melt(radar_urban_test, id.vars = 'group')

#Creating the radar plot
base  = ggplot(urban_toplot, aes(x = variable, y = value, group = group))

plot1 = base + geom_polygon(fill = 'gray',col='orange') + coord_polar()

plot2 = plot1 + facet_wrap(~group,
                           ncol = 3) 
plot2
```

